---
output: html_document
---
Exploratory Data Analysis of a "White Wine" data set by Manuhar Chopra
========================================================


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(reshape2)
library(dplyr)
library(RColorBrewer)
library(gridExtra)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('wineQualityWhites.csv')

```

This data set consists of a number of results from objective tests 
(e.g. pH,alcohol,density) on White wines. It also includes a quality rating, 
which is based on sensory data (median of at least 3 evaluations made by 
wine experts) between 0 (very bad) and 10 (very excellent). 

The citation of the data set is as below

# Citation

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
Modeling wine preferences by data mining from physicochemical properties.
In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

Available at: 

          [@Elsevier] http://dx.doi.org/10.1016/j.dss.2009.05.016
          
          [Pre-press (pdf)] http://www3.dsi.uminho.pt/pcortez/winequality09.pdf
          
          [bib] http://www3.dsi.uminho.pt/pcortez/dss09.bib


# Univariate Plots Section

First, performing some preliminary exploration of the data set to understand 
the structure and view the summaries of different variables

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots}
str(df)
summary(df)
```

In the data set, the mean quality is 5.878, with a median quality of 6. 
Density and pH appear to be less dispersed than other variables, with
their mean & median rather close.

To better visualize the distribution of quality, a bar chart is plotted below,
with the color also reflecting the quality for better visualization. Also, 
creating a new variable (quality.f), which comes from factoring the original 
quality variable. This transformation may be useful in the future as it is now
a discrete variable

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#creating a new variable from factoring quality
df$quality.f<-factor(df$quality,ordered = TRUE)

#plotting a bar chart
ggplot(data=df,aes(x=quality.f,fill=quality.f))+
  geom_bar()+
  scale_fill_brewer(type='seq')

summary(df$quality)
```

Matching with our expectations, 6 has the highest count of wines. 

Further analyzing the variables, it can be seen that both total and free sulfur
dioxide are present in the data set. However, it seems that "bound"
sulfur dioxide (total-free) is missing. Hence, this additional variable 
is created below, such that 

bound.sulfur.dioxide = total.sulfur.dioxide-free.sulfur.dioxide

The summary of this new variable is as below

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#creating a new variable based on two existing variables
df$bound.sulfur.dioxide = with(df,total.sulfur.dioxide-free.sulfur.dioxide)

summary(df$bound.sulfur.dioxide)
```

Now, to visually explore each of the other variables, they were plotted using
grid arrange

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#creating plots
h1 <- ggplot(aes(x = fixed.acidity), data = df) + 
  geom_histogram()
h2 <- ggplot(aes(x = volatile.acidity), data = df) +
  geom_histogram()
h3 <- ggplot(aes(x = citric.acid), data = df) +
  geom_histogram()
h4 <- ggplot(aes(x = residual.sugar), data = df) +
  geom_histogram()
h5 <- ggplot(aes(x = chlorides), data = df) +
  geom_histogram()
h6 <- ggplot(aes(x = free.sulfur.dioxide), data = df) +
  geom_histogram()
h7 <- ggplot(aes(x = total.sulfur.dioxide), data = df) +
  geom_histogram()
h8 <- ggplot(aes(x = bound.sulfur.dioxide), data = df) +
  geom_histogram()
h9 <- ggplot(aes(x = density), data = df) +
  geom_histogram()
h10 <- ggplot(aes(x = pH), data = df) +
  geom_histogram()
h11 <- ggplot(aes(x = sulphates), data = df) +
  geom_histogram()
h12 <- ggplot(aes(x = alcohol), data = df) +
  geom_histogram()

#printing the plots together
grid.arrange(h1, h2, h3, h4, h5, h6, h7, h8, h9, h10, h11, h12)
```

Most of the plots seem to roughly follow a normal distribution, with the
exception of residual sugar. Perhaps transforming it
with the log function will result in a better trans formation.

Trying out the log transform on the residual sugar variable below to address 
the long tail

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = residual.sugar), data = df) +
  geom_histogram(binwidth = .2) +
  scale_x_continuous(trans = "log")
```

The residual sugar histogram now shows a bi-modal pattern. This data might be
useful in further analysis in the following sections

Checking below to see how many wines have total SO2 > 50, which is the 
threshold value when it starts becoming noticeable in the taste

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#Ratio of high SO2 to total wines
nrow(df[df$total.sulfur.dioxide>50,])/nrow(df)
```

Most of the wines (99%) in this database have a high SO2 content, which should
be noticeable in their taste.Lets see in the further sections if this is valid

For most wines, pH should between 3 and 4. Lets see below how many wines fall 
outside this expected value range of pH

```{r,echo=FALSE, message=FALSE, warning=FALSE}

#Ratio of non standard pH to total wines
nrow(df[df$pH>4 | df$pH<3,])/nrow(df)
```

Less than .01% lie outside the 3-4 range, with the most clustered between 2.9 
and 3.

However, there seem to be some outliers in the data. Outliers can reduce the
robustness of data analysis and can induce bias.Representing outliers with
the graphs below

```{r,echo=FALSE, message=FALSE, warning=FALSE}
h1 <- ggplot(aes(x = 1, y = fixed.acidity), data = df) +
    geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red') 
h2 <- ggplot(aes(x = 1, y = volatile.acidity), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red')
h3 <- ggplot(aes(x = 1, y = citric.acid), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red') 
h4 <- ggplot(aes(x = 1, y = log(residual.sugar)), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red')
h5 <- ggplot(aes(x = 1, y = chlorides), data = df) +
   geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h6 <- ggplot(aes(x = 1, y = free.sulfur.dioxide), data = df) +
   geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h7 <- ggplot(aes(x = 1, y = total.sulfur.dioxide), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h8 <- ggplot(aes(x = 1, y = bound.sulfur.dioxide), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h9 <- ggplot(aes(x = 1, y = density), data = df) +
   geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h10 <- ggplot(aes(x = 1, y = pH), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h11 <- ggplot(aes(x = 1, y = sulphates), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
h12 <- ggplot(aes(x = 1, y = alcohol), data = df) +
  geom_jitter(alpha = 0.1 ) + geom_boxplot(alpha = 0.2, color = 'red' )
#Display all the plots 
grid.arrange(h1, h2, h3, h4, h5, h6, h7, h8, h9, h10, h11, h12)
```

Although outliers can be clearly seen, they are not deleted from the data set
as they could potentially hold some insights into the data.

Creating a final iteration of the grid of the variables,
not plotting the outliers (bottom & top 1%) for all variables


```{r,echo=FALSE, message=FALSE, warning=FALSE}
h1 <- ggplot(aes(x = fixed.acidity), data = df) +
  xlim(quantile(df$fixed.acidity, 0.01),
       quantile(df$fixed.acidity, 0.99)) +
  geom_histogram()
h2 <- ggplot(aes(x = volatile.acidity), data = df) +
  xlim(quantile(df$volatile.acidity, 0.01),
       quantile(df$volatile.acidity, 0.99)) +
  geom_histogram()
h3 <- ggplot(aes(x = citric.acid), data = df) +
  xlim(quantile(df$citric.acid, 0.01),
       quantile(df$citric.acid, 0.99)) +
  geom_histogram()
h4 <- ggplot(aes(x = residual.sugar), data = df) +
  xlim(quantile(df$residual.sugar, 0.01),
       quantile(df$residual.sugar, 0.99)) +
  geom_histogram(binwidth = .2) +
  scale_x_continuous(trans = "log")
h5 <- ggplot(aes(x = chlorides), data = df) +
  xlim(quantile(df$chlorides, 0.01),
       quantile(df$chlorides, 0.99)) +
  geom_histogram()
h6 <- ggplot(aes(x = free.sulfur.dioxide), data = df) +
  xlim(quantile(df$free.sulfur.dioxide, 0.01),
       quantile(df$free.sulfur.dioxide, 0.99)) +
  geom_histogram()
h7 <- ggplot(aes(x = total.sulfur.dioxide), data = df) +
  xlim(quantile(df$total.sulfur.dioxide, 0.01),
       quantile(df$total.sulfur.dioxide, 0.99)) +
  geom_histogram()
h8 <- ggplot(aes(x = bound.sulfur.dioxide), data = df) +
  geom_histogram()
h9 <- ggplot(aes(x = density), data = df) +
  xlim(quantile(df$density, 0.01),
       quantile(df$density, 0.99)) +
  geom_histogram()
h10 <- ggplot(aes(x = pH), data = df) +
  xlim(quantile(df$pH, 0.01),
       quantile(df$pH, 0.99)) +
  geom_histogram()
h11 <- ggplot(aes(x = sulphates), data = df) +
  xlim(quantile(df$sulphates, 0.01),
       quantile(df$sulphates, 0.99)) +
  geom_histogram()
h12 <- ggplot(aes(x = alcohol), data = df) +
  xlim(quantile(df$alcohol, 0.01),
       quantile(df$alcohol, 0.99)) +
  geom_histogram(binwidth = .2)

#Display all the plots 
grid.arrange(h1, h2, h3, h4, h5, h6, h7, h8, h9, h10, h11, h12)

```

# Univariate Analysis

### What is the structure of your dataset?
There are 11 quantitative features and 4,898 observations in the data set.

The final variable is quality which was converted to a factor, since it 
provides a rating from 0 to 10.

The actual quality varies from 3 to 9, with 6 as median and  5.88 as mean.


### What is/are the main feature(s) of interest in your dataset?
A useful final model should be able to predict the "quality" of the wine. 
Hence, that is the key feature of interest. 

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
Many of the other features likely have an impact on the final quality.
For example alcohol,density, pH appear to be of interest

### Did you create any new variables from existing variables in the dataset?
As the data only included free and total sulfur dioxide (sum of free & bound),
I created a new variable - 
"bound sulfur dioxide" = Total - Free Sulfur Di oxide.

This variable may provide additional insights going forward in the analysis.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I made the following changes in the data set

1. Added a new variable, quality.f, which is a factor of the original 
quality variable, as the ratings can be better analyzed as discrete values.

2. The sugar and alcohol data did not appear to be normally distributed among
all the other variables.Hence, I used the Log transformation on the 
residual sugar and alcohol values. Thus they became closer to a normal 
distribution


# Bivariate Plots Section

As a first step, the ggpairs function was used to get an overview of the 
various relationships existing between pairs of variables. The X and the 
quality.f variables were excluded to have a cleaner analysis.

Some of the variable names have been shortened to increase readability,
e.g. b.SO2 = bound sulfur dioxide


```{r,echo=FALSE, message=FALSE, warning=FALSE}

str(df)
df.s<-df
names(df.s)[names(df.s) == 'fixed.acidity'] <- 'f.acid'
names(df.s)[names(df.s) == 'volatile.acidity'] <- 'v.acid'
names(df.s)[names(df.s) == 'citric.acid'] <- 'c.acid'
names(df.s)[names(df.s) == 'residual.sugar'] <- 'sugar'
names(df.s)[names(df.s) == 'free.sulfur.dioxide'] <- 'f.SO2'
names(df.s)[names(df.s) == 'bound.sulfur.dioxide'] <- 'b.SO2'
names(df.s)[names(df.s) == 'total.sulfur.dioxide'] <- 't.SO2'

ggcorr(df.s[, 2:15], label = TRUE,hjust = .6, size = 3)
   

```

The highest correlations with quality (positive or negative) which came from
the above result are listed below

Quality and alcohol: 0.436

Quality and density: -0.307

Quality and Bound Sulfur Dioxide: -0.218

Quality and chlorides: -0.210

Quality and Total Sulfur Di Oxide :-.175

None of the values point towards a strong correlation !

In order to further analyse these variables, I am plotting below a scatter plot
with for alcohol vs Quality. Including jitter, alpha & removing 1% outliers
on both ends for better visualization.

Additionally, adding a geom_smooth overlay on the data to get a sense of
the variation

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = quality), data = df) +
  geom_jitter(alpha=0.05)+
  xlim(quantile(df$alcohol, 0.01), quantile(df$alcohol, 0.99))+
  geom_smooth()

cor(df$quality,df$alcohol)
```

The o.436 correlation can be seen, as higher % of alcohol is leading to a 
higher quality score within our given range.

However, there seems to be a slight dip in quality around 9-10 % alcohol. Since
alcohol has the strongest correlation, analyzing this dip in detail in the plot
below. The X & Y axis have been interchanged, and a stat_summary line plots
the mean alcohol % for a given quality

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(y = alcohol, x = quality.f), data = df) +
 geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 7, 
               size = 4)+
  ylim(quantile(df$alcohol, 0.01), quantile(df$alcohol, 0.99))
```

It appears there is an inflection point at quality =5, where the line changes 
slope. I.e. this means that for lower quality wine (<5), the correlation is
negative and for higher quality wine, the correlation is positive.

Trying to find these correlation factors below

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#finding correlation for the two ranges of quality
low_quality_cor<- cor(df[df$quality<5,]$quality,df[df$quality<5,]$alcohol)
high_quality_cor<- cor(df[df$quality>=5,]$quality,
                            df[df$quality>=5,]$alcohol)
c(low_quality_cor,high_quality_cor)
```

Hence, as it was suspected, the correlation between alcohol & quality is 
-0.0586 if quality is less than 5.
If the quality is equal or more than 5, then the correlation is 0.467. This
means that a simple linear model will not be very accurate. A conditional or
non linear model may be more suitable.


Plotting similar comparison histograms for density, bound sulfur dioxide, 
chlorides and total sulfur dioxide below


```{r,echo=FALSE, message=FALSE, warning=FALSE}
p1<-ggplot(aes(y = density, x = quality.f), data = df)+
  geom_jitter(alpha=0.05)+
  ylim(quantile(df$density, 0.01),quantile(df$density, 0.99))+
  geom_boxplot(alpha=.3,color='blue')+
  stat_summary(fun.y = "mean",geom = "point",color = "red",shape = 7,size = 4)

p2 <- ggplot(aes(y = bound.sulfur.dioxide, x = quality.f), data = df) +
  geom_jitter(alpha=0.05)+
  ylim(quantile(df$bound.sulfur.dioxide, 0.01), 
       quantile(df$bound.sulfur.dioxide, 0.99))+
  geom_boxplot(alpha=.3,color='blue')+
  stat_summary(fun.y = "mean",geom = "point",color = "red",shape = 7,size = 4)

p3 <- ggplot(aes(y = chlorides, x = quality.f), data = df) +
  geom_jitter(alpha=0.05)+
  ylim(quantile(df$chlorides, 0.05), quantile(df$chlorides, 0.95))+
  geom_boxplot(alpha=.3,color='blue')+
  stat_summary(fun.y = "mean",geom = "point",color = "red",shape = 7,size = 4)

p4 <- ggplot(aes(y = total.sulfur.dioxide, x = quality.f), data = df) +
  geom_jitter(alpha=0.05)+
  ylim(quantile(df$total.sulfur.dioxide, 0.01), 
       quantile(df$total.sulfur.dioxide, 0.99))+
  geom_boxplot(alpha=.3,color='blue')+
  stat_summary(fun.y = "mean",geom = "point",color = "red",shape = 7,size = 4)

grid.arrange(p1,p2,p3,p4)


```

The following were the key takeaways from the above charts

1.Density and quality have a weak negative correlation (-0.3), 
with higher densities associated with lower quality

2.Bound Sulfur Dioxide (-0.22)& Chlorides (-0.21) have very weak 
negative correlations, with quality generally falling as values of these 
variables increase

3.Total Sulfur Dioxide and quality has the weakest negative 
correlation (-0.175) among the pairs which we selected. It is weak enough 
to not be able to draw conclusive results

However, a point to note is that for both bound & total sulfur dioxide, there
is a weak positive correlation which then turns negative at around 5 quality
level.This could be because at that levels it starts to negatively impact
taste

The other variable pairs which had a high correlation among them are as below

Total sulfur dioxide and bound sulfur dioxide: 0.922

Density and residual sugar: 0.839

Alcohol and density: -0.780

Total sulfur dioxide and free sulfur dioxide: 0.616

Bound sulfur dioxide and density: 0.504

Clearly, sulfur-dioxide levels (total vs bound or total vs free) will be
strongly correlated to each other.
Leaving those aside , we get Density VS Alcohol, Density VS sugar & 
Density VS Bound SO2

```{r,echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(aes(x = density, y = alcohol), data = df) +
  geom_jitter(alpha = .1)+
  xlim(quantile(df$density, 0.01), quantile(df$density, 0.99)) +
  geom_smooth(method = "lm", se = FALSE,size=1.5)

p2 <- ggplot(aes(x = density, y = residual.sugar), data = df) +
  geom_jitter(alpha = .1) +
  xlim(quantile(df$density, 0.01), quantile(df$density, 0.99)) +
  geom_smooth(method = "lm", se = FALSE,size=1.5)+
  ylim(quantile(df$residual.sugar, 0.01), quantile(df$residual.sugar, 0.99))

p3 <- ggplot(aes(x = density, y = bound.sulfur.dioxide), data = df) +
  geom_jitter(alpha = .1) +
  xlim(quantile(df$density, 0.01), quantile(df$density, 0.99)) +
  geom_smooth(method = "lm", se = FALSE,size=1.5)+
   ylim(quantile(df$bound.sulfur.dioxide, 0.01), 
        quantile(df$bound.sulfur.dioxide, 0.99))

grid.arrange(p1,p2,p3)

```

Alcohol and density have a strong negative correlation (-0.780). This may 
seem logical as alcohol is volatile liquid with a low density. Higher alcohol % 
would then tend to reduce the overall density of the wine

Sugar, has a bit the reverse effect of alcohol. It is strongly positively 
correlated (0.839) with density. This is probably as sugar itself is 
rather dense, its higher amounts increases the overall wine density

Bound SO2 has a largely positive correlation (0.504), though not as 
strong as sugar. 


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data set?

The feature of interest (quality) has weak correlation with alcohol (0.436),
density (-.307),bound sulfur dioxide (-0.218), chlorides(-0.210) & 
Total Sulfur dioxide variables. (-0.175)

1.With Alcohol increasing from 3 to 5, the quality decreased. As the alcohol 
increased from 5 to 9, the quality increased.

2.Quality weakly decreases with an increase in either of the remaining four
variables, i.e. density, bound sulfur dioxide,chlorides and total sulfur

However, a point to note is that for both bound & total sulfur dioxide, there
is initially a weak positive correlation which then turns negative around 70-80
ppm. This could be because at that levels it starts to negatively impact
taste

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

There were strong correlations between the following variables

Total sulfur dioxide and bound sulfur dioxide: 0.922

Density and residual sugar: 0.839

Alcohol and density: -0.780

Total sulfur dioxide and free sulfur dioxide: 0.616

Bound sulfur dioxide and density: 0.504

The relationships between Free, bound & total sulfur seem obvious. Besides
those density was strongly positively linked with residual sugar (sugar has 
high density) and negatively linked with alcohol (alcohol has low density).
Density was also weakly positively linked with sulfur dioxide


### What was the strongest relationship you found?
Excluding the correlation between total & bound sulfur (as they are related
variables), the best correlation was between density and residual sugar.

# Multivariate Plots Section

The last section brought to light some interesting relationships between the
various variable pairs. In this section, I will deep dive into some of those
relationships, adding adding data from additional variables to look for
further insights

Plotting below the alcohol vs residual sugar,where color depends on quality

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = residual.sugar  , color = quality.f), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$alcohol, 0.01), quantile(df$alcohol, 0.99))+
  ylim(quantile(df$residual.sugar, 0.01), quantile(df$residual.sugar, 0.99))+
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))+
  theme_dark()
```

We can see that residual sugar is higher at lower alcohol levels, and
lower at higher alcohol levels. This also fits with the understood logic
that the sugars are fermented to create alcohol in the process of wine making.
Also, higher quality wines tend to have higher amounts of alcohol

Plotting below to check if a similar relationship exists between quality,
density & residual sugar

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = density, y = residual.sugar  , color = quality.f), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$density, 0.01), quantile(df$density, 0.99))+
  ylim(quantile(df$residual.sugar, 0.01), quantile(df$residual.sugar, 0.99))+
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))+
  theme_dark()
```

The plot above reflects the findings of two of our earlier discussion points.
Quality is generally falling as density increases, and that higher densities
are associated with higher residual sugar levels

Plotting below to check if a similar relationship exists between quality,
bound sulfur dioxide and residual sugar

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = bound.sulfur.dioxide, y = residual.sugar  , color = quality.f), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$bound.sulfur.dioxide, 0.01), 
       quantile(df$bound.sulfur.dioxide, 0.99))+
  ylim(quantile(df$residual.sugar, 0.01), quantile(df$residual.sugar, 0.99))+
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))+
  theme_dark()
```
 
This graph shows that elevated levels of bound SO2 tend to be 
 positively correlated with higher amounts of residual sugar. This sounds 
 logical as sugar is known to bind with sulfur to create bound SO2.Also,
 better quality is generally associated lower levels of bound
 sulfur dioxide
 
 On the same topic, plotting the ratio of bound So2 / total SO2 with residual 
 sugar
 
```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = bound.sulfur.dioxide/total.sulfur.dioxide,
           y = residual.sugar  , color = quality.f), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$bound.sulfur.dioxide/df$total.sulfur.dioxide, 0.01), 
       quantile(df$bound.sulfur.dioxide/df$total.sulfur.dioxide, 0.99))+
  ylim(quantile(df$residual.sugar, 0.01), quantile(df$residual.sugar, 0.99))+
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))+
  theme_dark()
```

The graph shows some interesting results. At most low levels of quality,
its is showing a negative correlation between residual sugar and proportion
of bound So2. But at most high levels of quality, this correlation seems to
become positive ! As they say, correlation does not imply causation, but 
this is still a point to note

Leaving aside quality, the other highly correlated features were alcohol, 
density and residual sugar. To better understand their distribution, I have 
distributed these variables below into "buckets"

```{r,echo=FALSE, message=FALSE, warning=FALSE}

df$alcohol.bucket <- 
  with(df, cut(alcohol, breaks = c(7,9.5,10.4,11.4, 14.2),
                  ordered = TRUE))
df$density.bucket <- 
  with(df, cut(density, breaks = c(.9871,.9917,.9937,.9961,1.039),
                  ordered = TRUE))
df$chlorides.bucket <- 
  with(df, cut(chlorides, breaks = c(.008,.036,.043,.05,.346),
                  ordered = TRUE))
df$residual.sugar.bucket <- 
  with(df, cut(residual.sugar, breaks = c(.5,5.2,65.8),
                  ordered = TRUE))

ggplot(aes(x = alcohol, y = density  , color = residual.sugar.bucket), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$alcohol, 0.01), 
       quantile(df$alcohol, 0.99))+
  ylim(quantile(df$density, 0.01), quantile(df$density, 0.99))+
  scale_color_brewer(type='div', guide=guide_legend(title='Sugar'))+
  theme_dark()
```



Alcohol and density are negatively correlated, with high sugar contributing
to higher densities

Faceting this graph by the quality levels yields the result below

```{r,echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = alcohol, y = density  , color = residual.sugar.bucket), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$alcohol, 0.01), 
       quantile(df$alcohol, 0.99))+
  ylim(quantile(df$density, 0.01), quantile(df$density, 0.99))+
  scale_color_brewer(type='div', guide=guide_legend(title='Sugar'))+
  theme_dark()+
  facet_wrap(~quality.f)
```

High quality (quality >6) wines tend be more in the bottom right
corner, with higher alcohol and lower densities. Quality 9 wines are in fact 
mostly low sugar, low density and high alcohol !

A linear model was attempted from this data, to predict wine qualities. 
Starting with the highest correlation variable (alcohol), each of the other
variables were added one by one to select the best model

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#Building the model, adding one variable at a time
m1 <- lm(quality ~ alcohol, data = df)
m2 <- update(m1, ~ . + density)
m3 <- update(m2, ~ . + chlorides)
m4 <-update(m3, ~ . + pH)
m5 <- update(m3, ~ . + bound.sulfur.dioxide)
m6 <- update(m4, ~ . + fixed.acidity)
m7 <- update(m5, ~ . + total.sulfur.dioxide)
m8 <- update(m6, ~ . + residual.sugar)
m9 <- update(m7, ~ . + citric.acid)
m10 <- update(m8, ~ . + sulphates)
m11 <- update(m9, ~ . + free.sulfur.dioxide)
m12 <- update(m10, ~ . + volatile.acidity)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)
```

The highest R squared value achieved was 0.278, which does not amount to a
very robust model. Possible explanations could be that either the model is
not linear (which some of our analysis points towards) or that there are other
features which are affecting the final quality

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In this section, I observed the following relationships

1. Quality vs alcohol & Residual Sugar: The positive correlation between quality 
and alcohol of the earlier analysis was added another dimension by residual
sugar. Lower residual sugar tended to increase the alcohol content, and hence
increase the positive impact on quality

2. Quality vs density & Residual Sugar: Higher residual sugars increased the
density, which tended to lower the overall quality of the wine. Hence, residual
sugar & density strengthened their negative impact on the overall quality.

3. Quality vs bound sulfur dioxide & Residual Sugar: Higher residual sugars 
tended increased the amount of bound sulfur dioxide, which increased the
negative impact on the quality of the wine



### Were there any interesting or surprising interactions between features?

1. One surprising interaction was the one between bound sulfur dioxide and 
residual sugar. It seems higher sugar tends to bind more
Sulfur di oxide,negatively impacting the overall quality of the wine

2. The other interesting interaction was between alcohol, density and residual
sugar. Higher sugar correlates with higher density, while higher sugar also
correlates with lower alcohol.Given the individual correlations of each 
of the variables with quality, its makes sense that the best quality 
wines tends to have low residual sugar, low density & high alcohol.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

A linear model was fitted into the data set which attempts to predict the 
quality of the wine based on the the other input features.

As the number of input features was increased, the model generally improved,
except for certain variables, with the highest R squared value of 0.278 which 
includes all the variables

This value is not particularly high. As discussed before, this could be because 
the variable relationships are not linear or that there are some unknown
variables also affecting quality. Also, some input variables are internally
correlated to each other

------

# Final Plots and Summary

### Plot One
```{r ,echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

ggplot(aes(y = alcohol, x = quality.f), data = df) +
 geom_jitter( alpha = .3)  +
  geom_boxplot( alpha = .5,color = 'blue')+
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 7, 
               size = 4)+
  ylim(quantile(df$alcohol, 0.01), quantile(df$alcohol, 0.99))+
  xlab("Wine Quality") +
  ylab("Alcohol % ") +
  ggtitle(" Effect of Alcohol % on Wine Quality")
```

### Description One
This plot shows the important relationship between alcohol & wine quality. 
Although it has a positive correlation of 0.436, there seems 
to be a slight dip in quality around 9-10 % alcohol. 

It appears there is an inflection point at quality =5, where the line changes 
slope. I.e. this means that for lower quality wine (<5), the correlation is
negative and for higher quality wine, the correlation is positive.
This also means that a simple linear model will not be very accurate in
predicting this behavior. A conditional/non linear model may be more suitable.


### Plot Two
```{r ,echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(aes(x = bound.sulfur.dioxide, y = residual.sugar  , color = quality.f), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$bound.sulfur.dioxide, 0.01), 
       quantile(df$bound.sulfur.dioxide, 0.99))+
  ylim(quantile(df$residual.sugar, 0.01), quantile(df$residual.sugar, 0.99))+
  scale_color_brewer(type='seq', guide=guide_legend(title='Quality'))+
  theme_dark()+
  ylab("Residual Sugar gm/L") +
  xlab("Bound Sulphur Dioxide ") +
  ggtitle(" Effect of Residual Sugar on Bound SO2 ")
```

### Description Two
This scatter plot shows the bound sulfur dioxide vs 
residual sugar & quality.High residual sugar tends to increase 
the amount of bound sulfur dioxide, as higher sugar tends to covert
free Sulfur di oxide into bound sulfur dioxide.

High sugar and consequently higher bound SO2 negatively impacts the 
quality of the wine, which can be seen in the graph. Higher quality wines
generally tend to have lower bound SO2, with none of the highest quality (9)
wines having bound SO2> ~ 120

### Plot Three
```{r ,echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(aes(x = alcohol, y = density  , color = residual.sugar.bucket), 
       data = df) +
      geom_point(alpha = .1, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1.5)  +
  xlim(quantile(df$alcohol, 0.01), 
       quantile(df$alcohol, 0.99))+
  ylim(quantile(df$density, 0.01), quantile(df$density, 0.99))+
  scale_color_brewer(type='div', guide=guide_legend(title='Sugar'))+
  theme_dark()+
  xlab("Alcohol %") +
  ylab("Density gm/cm^3 ") +
  ggtitle(" Residual Sugar VS Density VS Alcohol % ")
  
```

### Description Three
This scatter plots shows Alcohol % VS density, with the color associated
with the level of residual sugar

Alcohol and density are negatively correlated. In addition, 
as sugar has itself a high density, higher amounts of sugar tend to increase
the overall density of the wine.

------

# Reflection
This project has helped increase both my knowledge of exploratory data analysis
using R and of white wines !

This database contained 4898 data points about white wines, with each data point
associated with 11 characteristics such as residual sugar, density, 
alcohol etc along with an overall wine quality score

I started off with getting to understand the data and exploring the given
variables through individual variable plots. I then plotted variables
against each other and the overall wine quality to better understand the 
nature and amount of their impact.Finally, selecting those variables which 
had the maximum impact, I performed advance plotting, visualizing and 
understanding multiple variables in each plot to draw interesting insights
from the data

Certain variables like alcohol had a positive correlation with quality, while
certain variables like density had a negative correlation with quality.
It was interesting that many of these variables were also affecting each
other, e.g. how alcohol and density are negatively correlated. I also created a 
new variable "bound sulfur dioxide" which eventually uncovered an interesting 
relationship with residual sugar.

In the end, I  created a linear model to try and predict the quality of the wine
based on the wine characteristics. The overall robustness of this model did not
come out to be very high. I tried to solve this by adding one variable
at a time, to try and find a combination which fits best. This process
somewhat improved the R squared value, but was still not high enough to
"Pop the Champagne" ! Perhaps opening some regular wine would suffice !

This could mean that either the model is not linear or there are other 
unknown variables affecting the final quality.

It is important to remember that correlation does not imply causation ! Hence,
drawing conclusions from correlation alone is not recommended although it can
give a starting point from where further analysis can be undertaken. In this
project, further steps could be taken to try to fit a non linear model or 
explore if there are any other unknown variables present.

Nonetheless, it was a learning experience where I picked up knowledge of both
the concepts and the tools of Exploratory Data Analysis
